D-Hammer: Efficient Equational Reasoning for Labelled Dirac Notation

### 2025-01-31 03labelled.tex
update

The no-cloning theorem, a fundamental property of quantum computing, prevents us from copying an unknown quantum state. This requires an additional check on the valid registers--they should not include repeated quantum variables--which is often handled in programming languages via linear types. As such, we define the \emph{order-free} variable set of a register inductively by: 1) $\var(R) \triangleq \{r\}$, if $R\equiv r$ and 2) $\var(R) \triangleq \var(R_1) \cup\var(R_2)$, if $R\equiv (R_1,R_2)$.
We use the variable set to establish side conditions of typing valid registers and employ it as the type parameter for annotating labelled Dirac terms.

@@

\paragraph*{Typing rules.}
There are various rules for computing types and checking vadility of registers and labelled terms. Here we display some of the rules and see Appendix for all the rest.
\begin{gather*}
  \frac{
      E \vdash R : \reg(\sigma) \quad
      E \vdash Q : \reg(\tau)
      \quad \var(R) \cap \var(Q) = \emptyset
  }{E \vdash (R,Q) : \reg(\sigma \times \tau)} \\[0.2cm]
  \frac{
          E \vdash r : \reg(\sigma) \quad
          E \vdash i : \Basis(\sigma)}
  {E \vdash |i\>_r : \DType(\{r\}, \emptyset)}
  \qquad
  \frac{E \vdash R : \reg(\sigma)\qquad E \vdash K : \KType(\sigma)}{E \vdash K_R : \DType(\var(R), \emptyset)} \\[0.2cm]
    \frac{
        E \vdash D_i : \DType(s_i,s_i') \qquad
        \forall\,i\neq j.\ s_i\cap s_j = \emptyset \qquad
        \forall\,i\neq j.\ s'_i\cap s'_j = \emptyset
    }
    {E \vdash D_1 \otimes \cdots \otimes D_i : \DType(\bigcup_i s_i, \bigcup_i s_i')}.
\end{gather*}
The first rule says that a paired register is of the product type and its components should be disjoint.
To lift a plain Dirac notation into the labelled one (line 2), we enforce that the term and register are of the same indices, reflecting the fact the state should be consistant to the corresponding subsystems.
The third line gives the typing of labelled tensoring with a check whether the component subsystems are disjoint with each other.

@@

\begin{definition}[Cylindrical Extension]
  For any $D : \cD(s_1,s_2)$ and $s$ that disjoint with both $s_1$ and $s_2$, we define $\cl(D,s) \triangleq D\otimes \mathbf{1}_{s}$ of type $\cD(s_1\cup s, s_2\cup s)$.
\end{definition}
Formally, we equip $\cR$ with a default order, say for example the alphabetical order of names. For any valid register $R$, there exists the operator $\Swap_R$ that sorts $R$; for example, if $E\vdash p : \reg(\sigma_p)$ and $E\vdash q : \reg(\sigma_q)$, then $\Swap_{(q,p)} \triangleq \sum_{i\in \bU(\sigma_p)}\sum_{j\in \bU(\sigma_q)}|i\>\<j|\otimes |j\>\<i|$. We can further define $\Swap_{s_1,s_2,\cdots}$ for merging disjoint sets orderly.

For given context $E$ and set $s = \{r_1,\cdots,r_n\}\subseteq\cR$ ($r_i$ is sorted) with $E\vdash r_i : \sigma_i$, we define $\sigma_s\triangleq \sigma_1\times\cdots\times\sigma_n$ if $s$ is non-empty. \label{context: sigma s}
For any valuation $v$, we further write $\cH_{\sem{s}_v} \triangleq \bigotimes_{i}\cH_{\sem{\sigma_i}_v}$ which equals to $\cH_{\sem{\sigma_s}_v}$ for non-empty $s$.
% for any $s\subseteq \cR$, where $E\vdash r : \reg(\sigma_r)$ for all $r\in\cR$. 
We interpret labelled Dirac notations inductively as (assume $E\vdash D_i : \cD(s_i,s'_i)$) :
\begin{itemize}
  \item $\sem{|i\>_r}_v = |\sem{i}_v\>$;\quad 
        $\sem{\<i|_r}_v = \<\sem{i}_v|$;\quad  
        $\sem{K_R}_v = \sem{\Swap_R}_v \cdot \sem{K}_v$; \\
        $\sem{B_R}_v = \sem{B}_v\cdot \sem{\Swap_R}_v$;\quad
        $\sem{O_{R_1,R_2}} = \sem{\Swap_{R_1}}_v\cdot \sem{O}_v\cdot \sem{\Swap_{R_2}}_v$;
  \item $\sem{D_1\otimes \cdots \otimes D_n}_v = \sem{\Swap_{s_1,\cdot,s_n}}_v\cdot (\sem{D_1}_v\otimes \cdots \otimes \sem{D_1}_v)\cdot \sem{\Swap_{s'_1,\cdot,s'_n}}_v$;
  \item $\sem{D_1\cdot D_2}_v = \sem{\cl(D_1, s_2\backslash s'_1)}_v\cdot \sem{\cl(D_2, s'_1\backslash s_2)}_v$. Note that $s_2\backslash s'_1$ and $s'_1\backslash s_2$ are the minimal extension that make it interpretable. E.g., to interpret ${}_p\<i|\cdot |j\>_{p,q}$, we at least need to extend ${}_p\<i|$ to ${}_p\<i|\otimes I_{q}$.
\end{itemize}
Thus, $\sem{D}_v$ is a linear map from $\cH_{\sem{s'}_v}$ to $\cH_{\sem{s}_v}$ given $E\vdash D : \cD(s,s')$. Following the semantics, labelled tensor is independent of its order, i.e., $\sem{D_1\otimes D_2}_v = \sem{D_2\otimes D_1}_v$ and $\sem{D_1\otimes (D_2\otimes D_3)}_v = \sem{(D_1\otimes D_2)\otimes D_3}_v$, which ensures the soundness of treating labelled tensor as an AC symbol.

@@

\paragraph*{2. Rewriting to the normal form.}
We add three types of rules for dealing with operators on labeled terms, 1) recursively applying them to subterms, 2) push big operators out and 3) eliminate generalized composition and bra-ket pairs.
%Other symbols on labelled Dirac notation are also 
Take the rule for conjugate $(D_1 \cdot D_2)^\dagger \ \reduce\ D_2^\dagger \cdot D_1^\dagger$ as an example of 1).
For 2), we use distributivity rules for scaling, labelled tensor and generalized composition. For example, rule
\[
% \textrm{(R-SUM-PUSHD0)}
X_1 \otimes \cdots (\sum_{i \in M} D) \cdots \otimes X_2\ \reduce\ \sum_{i \in M} (X_1 \otimes \cdots D \cdots \otimes X_n)
\]
will lift summation to the outside.
Extra rules for 3) are established including:
% The final step operates on sum and dot product. They will lift summation to the outside, and eliminate the bra-ket pairs whenever possible.
\begin{align*}
    % & \textrm{(R-SUM-PUSHD0)}
    % && X_1 \otimes \cdots (\sum_{i \in M} D) \cdots \otimes X_2\ \reduce\ \sum_{i \in M} (X_1 \otimes \cdots D \cdots \otimes X_n) \\
    % %
    % & \textrm{(R-SUM-PUSHD1)}
    % && (\sum_{i \in M} D_1) \cdot D_2 \ \reduce\ \sum_{i \in M} (D_1 \cdot D_2) \\
    %
    & \textrm{(R-L-SORT0)}
    && A : \DType(s_1, s_2), B : \DType(s_1', s_2'), s_2 \cap s_1'=\emptyset \Rightarrow A \cdot B \ \reduce\ A \otimes B \\
    %
    & \textrm{(R-L-SORT1)}
    && {}_r\bra{i}\cdot\ket{j}_r \ \reduce\ \delta_{i, j} \\
    %
    & \textrm{(R-L-SORT2)}
    && {}_r\bra{i}\cdot(Y_1 \otimes \cdots \otimes \ket{j}_r \otimes \cdots \otimes Y_m) \ \reduce\ \delta_{i, j}.(Y_1  \otimes \cdots \otimes Y_m)
    %
\end{align*}
% These rules are added to the rewriting system in~\Cref{sec: decide} and executed together.
Assuming no variables of $\DType(s_1, s_2)$, 
repeating the application of the above rules yield the normal form of both sides of the equality--the addition of big operators: 
\begin{equation}
  \label{eqn:Labelled normal}
  \sum_{i}\cdots\sum_{j} a_1 . (\ket{i}_{p} \otimes \cdots \otimes \bra{j}_{q})
  + \cdots +
  \sum_{k}\cdots\sum_{l} a_m . (\ket{k}_{r} \otimes \cdots \otimes \bra{l}_{s})
\end{equation}
where each sum body is a tensor of labelled basis with scalar coefficients in plain Dirac notation.
\paragraph*{3. Ordering and elimination of quantum variables.} 
We further sort tensors of labelled basis in every sum body of Eqn. (\ref{eqn:Labelled normal}) by 1. ket first and 2. the default order of variables. This yield the same shape of every subterm on both sides of the equality, e.g., 
\begin{equation*}
  % \label{eqn:Labelled normal}
  \sum_{i}\cdots\sum_{j}\cdots \sum_{i'}\cdots\sum_{j'} a_1 . ((\ket{i}_p \otimes \cdots \otimes \ket{j}_q) \otimes (\bra{i'}_{p'}\otimes\cdots\otimes \bra{j'}_{q'}))
\end{equation*}
and thus it is equivalent to decide the equivalence of additions of subterms of:
\begin{equation}
  \label{eqn:Labelled normal1}
  \sum_{i}\cdots\sum_{j}\cdots \sum_{i'}\cdots\sum_{j'} a_1 . ((\ket{i} \otimes \cdots \otimes \ket{j}) \otimes (\bra{i'}\otimes\cdots\otimes \bra{j'}))
\end{equation}
which do not involve any labels.

Define $T_{s,s'}$ as $\SType$ if $s=s'=\emptyset$, $\KType(\sigma_s)$ if $s'=\emptyset$, $\BType(\sigma_{s'})$ if $s=\emptyset$ and $\OType(\sigma_s,\sigma_{s'})$ otherwise (see Sec. \ref{context: sigma s}). We claim the soundness of our normalization as:
% Then if the original equality is $E\vdash D_1 = D_2 : \cD(s,s')$, what remains to be check $E\vdash e_1 = e_2 : T_{s,s'}$.

\begin{theorem}[Soundness and Completeness of Normalization]
  Assuming $E\vdash D_1 : \cD(s,s')$, $E\vdash D_2 : \cD(s,s')$, 
  and no variables of $\DType(\cdot,\cdot)$ in $D_1,D_2$.
  Let $e_1 = e_2$ be obtained by above normalization procedure on $D_1 = D_2$. Then:
  \begin{itemize}
    \item (Soundness) $E\vdash e_1 = e_2 : T_{s,s'}$ implies $E\vDash D_1 = D_2$;
    \item (Completeness) $E\vDash D_1 = D_2$ implies $E\vDash e_1 = e_2$
  \end{itemize}
  %All rewriting rules for labelled Dirac notations listed here and in Appendix are sound, i.e., for any interpretation $\sem{\cdot}$, $D\reduce D'$ implies $\sem{D} = \sem{D'}$.
\end{theorem}
\begin{proof}
  Notice that all rewriting rules and ordering perserve the types, so the final step eliminating labels is both sound and complete. 
  It is routinely to check that each rewriting rules and ordering step are sound based on the semantics. This directly leads to the soundness. For completeness, we need further check if the normalization procedure always provide $e_1 = e_2$, i.e., it never gets stuck and terminates. Step 1 is trivial. For step 2, it can be further checked that 1. there is no circular rules and thus terminates, and 2. if it is not in form (\ref{eqn:Labelled normal}), there must exists one of the rules can be further applied. For step 3, ordering is always applicable. This completes the proof.
\end{proof}

>>>>

The no-cloning theorem, a fundamental property of quantum computing, prevents us from copying an unknown quantum state. This requires an additional check on the valid registers--they should not include repeated quantum variables--which is often handled in programming languages via linear types. As such, we define the \emph{order-free} variable set of a register as all quantum variables appearing in the register.
We use the variable set to establish side conditions of typing valid registers and employ it as the type parameter for annotating labelled Dirac terms.

@@

\paragraph*{Typing rules.}
There are various rules for computing types and checking vadility of registers and labelled terms. Here we display some of the rules and see Appendix for all the rest.
\begin{gather*}
  \frac{
      \Gamma \vdash R : \reg(\sigma) \quad
      \Gamma \vdash Q : \reg(\tau)
      \quad \var(R) \cap \var(Q) = \emptyset
  }{\Gamma \vdash (R,Q) : \reg(\sigma \times \tau)} \\[0.2cm]
  \frac{
          \Gamma \vdash r : \reg(\sigma) \quad
          \Gamma \vdash i : \Basis(\sigma)}
  {\Gamma \vdash |i\>_r : \DType(\{r\}, \emptyset)}
  \qquad
  \frac{\Gamma \vdash R : \reg(\sigma)\qquad \Gamma \vdash K : \KType(\sigma)}{\Gamma \vdash K_R : \DType(\var(R), \emptyset)} \\[0.2cm]
    \frac{
        \Gamma \vdash D_i : \DType(s_i,s_i') \qquad
        \forall\,i\neq j.\ s_i\cap s_j = \emptyset \qquad
        \forall\,i\neq j.\ s'_i\cap s'_j = \emptyset
    }
    {\Gamma \vdash D_1 \otimes \cdots \otimes D_i : \DType(\bigcup_i s_i, \bigcup_i s_i')}.
\end{gather*}
The first rule says that a paired register is of the product type and its components should be disjoint.
To lift a plain Dirac notation into the labelled one (line 2), we enforce that the term and register are of the same indices, reflecting the fact the state should be consistant to the corresponding subsystems.
The third line gives the typing of labelled tensoring with a check whether the component subsystems are disjoint with each other.

@@

\begin{definition}[Cylindrical Extension]
  For any $D : \cD(s_1,s_2)$ and $s$ that disjoint with both $s_1$ and $s_2$, we define $\cl(D,s) \triangleq D\otimes \mathbf{1}_{s}$ of type $\cD(s_1\cup s, s_2\cup s)$.
\end{definition}
Formally, we equip $\cR$ with a default order, say for example the alphabetical order of names. For any valid register $R$, there exists the operator $\Swap_R$ that sorts $R$; for example, if $\Gamma\vdash p : \reg(\sigma_p)$ and $\Gamma\vdash q : \reg(\sigma_q)$, then $\Swap_{(q,p)} \triangleq \sum_{i\in \bU(\sigma_p)}\sum_{j\in \bU(\sigma_q)}|i\>\<j|\otimes |j\>\<i|$. We can further define $\Swap_{s_1,s_2,\cdots}$ for merging disjoint sets orderly.

For given context $\Gamma$ and set $s = \{r_1,\cdots,r_n\}\subseteq\cR$ ($r_i$ is sorted) with $\Gamma\vdash r_i : \sigma_i$, we define $\sigma_s\triangleq \sigma_1\times\cdots\times\sigma_n$ if $s$ is non-empty. \label{context: sigma s}
For any valuation $v$, we further write $\cH_{\sem{s}_v} \triangleq \bigotimes_{i}\cH_{\sem{\sigma_i}_v}$ which equals to $\cH_{\sem{\sigma_s}_v}$ for non-empty $s$.
% for any $s\subseteq \cR$, where $\Gamma\vdash r : \reg(\sigma_r)$ for all $r\in\cR$. 
We interpret labelled Dirac notations inductively as (assume $\Gamma\vdash D_i : \cD(s_i,s'_i)$) :
\begin{itemize}
  \item $\sem{|i\>_r}_v = |\sem{i}_v\>$;\quad 
        $\sem{\<i|_r}_v = \<\sem{i}_v|$;\quad  
        $\sem{K_R}_v = \sem{\Swap_R}_v \cdot \sem{K}_v$; \\
        $\sem{B_R}_v = \sem{B}_v\cdot \sem{\Swap_R}_v$;\quad
        $\sem{O_{R_1,R_2}} = \sem{\Swap_{R_1}}_v\cdot \sem{O}_v\cdot \sem{\Swap_{R_2}}_v$;
  \item $\sem{D_1\otimes \cdots \otimes D_n}_v = \sem{\Swap_{s_1,\cdot,s_n}}_v\cdot (\sem{D_1}_v\otimes \cdots \otimes \sem{D_1}_v)\cdot \sem{\Swap_{s'_1,\cdot,s'_n}}_v$;
  \item $\sem{D_1\cdot D_2}_v = \sem{\cl(D_1, s_2\backslash s'_1)}_v\cdot \sem{\cl(D_2, s'_1\backslash s_2)}_v$. Note that $s_2\backslash s'_1$ and $s'_1\backslash s_2$ are the minimal extension that make it interpretable. E.g., to interpret ${}_p\<i|\cdot |j\>_{p,q}$, we at least need to extend ${}_p\<i|$ to ${}_p\<i|\otimes I_{q}$.
\end{itemize}
Thus, $\sem{D}_v$ is a linear map from $\cH_{\sem{s'}_v}$ to $\cH_{\sem{s}_v}$ given $\Gamma\vdash D : \cD(s,s')$. Following the semantics, labelled tensor is independent of its order, i.e., $\sem{D_1\otimes D_2}_v = \sem{D_2\otimes D_1}_v$ and $\sem{D_1\otimes (D_2\otimes D_3)}_v = \sem{(D_1\otimes D_2)\otimes D_3}_v$, which ensures the soundness of treating labelled tensor as an AC symbol.

@@

\paragraph*{2. Rewriting to the normal form.}
We add three types of rules for dealing with operators on labeled terms, 1) recursively applying them to subterms, 2) push big operators out and 3) eliminate generalized composition and bra-ket pairs.
%Other symbols on labelled Dirac notation are also 
Take the rule for conjugate $(D_1 \cdot D_2)^\dagger \ \reduce\ D_2^\dagger \cdot D_1^\dagger$ as an example of 1).
For 2), we use distributivity rules for scaling, labelled tensor and generalized composition. For example, rule
\[
% \textrm{(R-SUM-PUSHD0)}
X_1 \otimes \cdots (\sum_{i \in M} D) \cdots \otimes X_2\ \reduce\ \sum_{i \in M} (X_1 \otimes \cdots D \cdots \otimes X_n)
\]
will lift summation to the outside.
Extra rules for 3) are established including:
% The final step operates on sum and dot product. They will lift summation to the outside, and eliminate the bra-ket pairs whenever possible.
\begin{align*}
    % & \textrm{(R-SUM-PUSHD0)}
    % && X_1 \otimes \cdots (\sum_{i \in M} D) \cdots \otimes X_2\ \reduce\ \sum_{i \in M} (X_1 \otimes \cdots D \cdots \otimes X_n) \\
    % %
    % & \textrm{(R-SUM-PUSHD1)}
    % && (\sum_{i \in M} D_1) \cdot D_2 \ \reduce\ \sum_{i \in M} (D_1 \cdot D_2) \\
    %
    & \textrm{(R-L-SORT0)}
    && A : \DType(s_1, s_2), B : \DType(s_1', s_2'), s_2 \cap s_1'=\emptyset \Rightarrow A \cdot B \ \reduce\ A \otimes B \\
    %
    & \textrm{(R-L-SORT1)}
    && {}_r\bra{i}\cdot\ket{j}_r \ \reduce\ \delta_{i, j} \\
    %
    & \textrm{(R-L-SORT2)}
    && {}_r\bra{i}\cdot(Y_1 \otimes \cdots \otimes \ket{j}_r \otimes \cdots \otimes Y_m) \ \reduce\ \delta_{i, j}.(Y_1  \otimes \cdots \otimes Y_m)
    %
\end{align*}
% These rules are added to the rewriting system in~\Cref{sec: decide} and executed together.
Assuming no variables of $\DType(s_1, s_2)$, 
repeating the application of the above rules yield the normal form of both sides of the equality--the addition of big operators: 
\begin{equation}
  \label{eqn:Labelled normal}
  \sum_{i}\cdots\sum_{j} a_1 . (\ket{i}_{p} \otimes \cdots \otimes \bra{j}_{q})
  + \cdots +
  \sum_{k}\cdots\sum_{l} a_m . (\ket{k}_{r} \otimes \cdots \otimes \bra{l}_{s})
\end{equation}
where each sum body is a tensor of labelled basis with scalar coefficients in plain Dirac notation.
\paragraph*{3. Ordering and elimination of quantum variables.} 
We further sort tensors of labelled basis in every sum body of Eqn. (\ref{eqn:Labelled normal}) by 1. ket first and 2. the default order of variables. This yield the same shape of every subterm on both sides of the equality, e.g., 
\begin{equation*}
  % \label{eqn:Labelled normal}
  \sum_{i}\cdots\sum_{j}\cdots \sum_{i'}\cdots\sum_{j'} a_1 . ((\ket{i}_p \otimes \cdots \otimes \ket{j}_q) \otimes (\bra{i'}_{p'}\otimes\cdots\otimes \bra{j'}_{q'}))
\end{equation*}
and thus it is equivalent to decide the equivalence of additions of subterms of:
\begin{equation*}
  \sum_{i}\cdots\sum_{j}\cdots \sum_{i'}\cdots\sum_{j'} a_1 . ((\ket{i} \otimes \cdots \otimes \ket{j}) \otimes (\bra{i'}\otimes\cdots\otimes \bra{j'}))
\end{equation*}
which do not involve any labels.

Define $T_{s,s'}$ as $\SType$ if $s=s'=\emptyset$, $\KType(\sigma_s)$ if $s'=\emptyset$, $\BType(\sigma_{s'})$ if $s=\emptyset$ and $\OType(\sigma_s,\sigma_{s'})$ otherwise (see Sec. \ref{context: sigma s}). We claim the soundness of our normalization as:
% Then if the original equality is $\Gamma\vdash D_1 = D_2 : \cD(s,s')$, what remains to be check $\Gamma\vdash e_1 = e_2 : T_{s,s'}$.

\begin{theorem}[Soundness and Completeness of Normalization]
  Assuming $\Gamma\vdash D_1 : \cD(s,s')$, $\Gamma\vdash D_2 : \cD(s,s')$, 
  and no variables of $\DType(\cdot,\cdot)$ in $D_1,D_2$.
  Let $e_1 = e_2$ be obtained by above normalization procedure on $D_1 = D_2$. Then:
  \begin{itemize}
    \item (Soundness) $\Gamma\vdash e_1 = e_2 : T_{s,s'}$ implies $\Gamma\vDash D_1 = D_2$;
    \item (Completeness) $\Gamma\vDash D_1 = D_2$ implies $\Gamma\vDash e_1 = e_2$
  \end{itemize}
  %All rewriting rules for labelled Dirac notations listed here and in Appendix are sound, i.e., for any interpretation $\sem{\cdot}$, $D\reduce D'$ implies $\sem{D} = \sem{D'}$.
\end{theorem}
% \begin{proof}
%   Notice that all rewriting rules and ordering perserve the types, so the final step eliminating labels is both sound and complete. 
%   It is routinely to check that each rewriting rules and ordering step are sound based on the semantics. This directly leads to the soundness. For completeness, we need further check if the normalization procedure always provide $e_1 = e_2$, i.e., it never gets stuck and terminates. Step 1 is trivial. For step 2, it can be further checked that 1. there is no circular rules and thus terminates, and 2. if it is not in form (\ref{eqn:Labelled normal}), there must exists one of the rules can be further applied. For step 3, ordering is always applicable. This completes the proof.
% \end{proof}
